The increasing cost of happiness

### 2020-09-10 drafts/draft0.Rmd
Added results

After adjusting for age, relationship status, and education level, we find that satisfaction increases linearly with household wealth, but the same purchasing power has diminishing returns on happiness. We also discover that the location of the inflection point has increased in real dollar value since 2002.

@@

#### Wealth
We have a variety of income & wealth information in HILDA, most of it collected annually. However the household net worth is collected every four years. In the present report we use:

1. [hwnwip](https://www.online.fbe.unimelb.edu.au/HILDAodd/KWCrossWaveCategoryDetails.aspx?varnt=hwnwip){target="_blank"} Household Net Worth ($) positive values [imputed] [weighted topcode]. Calculated as assets minus debts for the household.
2. [hifdip](https://www.online.fbe.unimelb.edu.au/HILDAodd/KWCrossWaveCategoryDetails.aspx?varnt=hifdip){target="_blank"} Household financial year disposable regular income ($) positive values [imputed] [weighted topcode]. The sum across all household members of financial year gross regular income minus taxes on financial year gross regular income.
3. [tifdip](https://www.online.fbe.unimelb.edu.au/HILDAodd/KWCrossWaveCategoryDetails.aspx?varnt=tifdip){target="_blank"} Personal disposable income ($) positive values [weighted topcode]. Financial year gross regular income minus taxes on financial year gross regular income.

_Note_. We also examined household and personal gross wages, and household and personal total disposable income.

Income and wealth variables were provided by the University of Melbourne as imputed, and with weighted topcodes to preserve anonymity. We removed individuals with values above the threshold indicated for top-coding.

Household wealth variables were adjusted for the number of people in each household in the same manner as Headey, Muffels & Wooden, 2004; i.e., income divided by the square root of household size (Buhmann et al, 1988; Coulter et al, 1992).

@@

@@

Age (and age^2^) and education were included as covariates, and students were removed.

Years of education were determined from the [Australian Qualifications Framework (AQF)](https://en.wikipedia.org/wiki/Australian_Qualifications_Framework#Certificates_I-IV){target="_blank"} from the Department of Industry.

Separate analyses for women are included in the Supplementary material.

@@

#### Demographics and sample characteristics

After stratification by decile (personal disposable income) we see that age, proportion of males, couples, years of education and education levels all increase with decile. Conversely, proportion of females, single people (never married), unemployed, and fulltime students decreases with decile. Of particular interest is that wellbeing, household wealth and SEIFA index follow a U-shaped function with the minimum value falling between the third and sixth decile.

@@

#### Household disposable income (adjusted for household size)

@@

as.data.frame(hifdip_gh9_fixed_fit) %>%
  select(b_alpha_Intercept, b_alpha_year2006, b_alpha_year2010, 
         b_alpha_year2014, b_alpha_year2018) %>%
  transmute(
    fy2002 = b_alpha_Intercept,
    fy2006 = mean(b_alpha_Intercept) + b_alpha_year2006,
    fy2010 = mean(b_alpha_Intercept) + b_alpha_year2010,
    fy2014 = mean(b_alpha_Intercept) + b_alpha_year2014,
    fy2018 = mean(b_alpha_Intercept) + b_alpha_year2018
  ) %>%
  mutate_all(.funs = ~brms::inv_logit_scaled(., 0, 150)*10) -> d1

color_scheme_set("red")
mcmc_areas(
  d1,
  pars = rev(colnames(d1)),
  prob = 0.95, # 80% intervals
  prob_outer = 1, # 99%
  point_est = "mean"
  ) +
  theme_test() +
  geom_vline(aes(xintercept = mean(d1$fy2002)), color = "gray", linetype = 3) +
  labs(
    subtitle = "Posterior change points (mean & 95% distribution)",
    x = "Household wealth ($000s)")
```

@@

- The change point for happiness (affective wellbeing) shifts to the right over time and requires greater household wealth.

@@

>>>>

After adjusting for age, gender, and education level, we find that satisfaction increases linearly with household wealth, but the same purchasing power has diminishing returns on happiness. We also discover that the location of the inflection point has increased in real dollar value since 2002.

@@

#### Wealth
We have a variety of income & wealth information in HILDA, most of it collected annually. However the household net worth is collected every four years. In the present report we use [hifdip](https://www.online.fbe.unimelb.edu.au/HILDAodd/KWCrossWaveCategoryDetails.aspx?varnt=hifdip){target="_blank"} Household financial year disposable regular income ($). The sum across all household members of financial year gross regular income minus taxes on financial year gross regular income. Wealth variables were provided by the University of Melbourne as imputed, and with weighted topcodes to preserve anonymity. _Nb_. We also examined household and personal gross wages and disposable income (not shown).

We removed individuals with values above the threshold indicated for top-coding. Household wealth was adjusted for the number of people in each household in the same manner as Headey, Muffels & Wooden, 2004; i.e., income divided by the square root of household size (Buhmann et al, 1988; Coulter et al, 1992).

@@

```{r funtimes}
source("../src/happyfun.r")
```

@@

Age (and age^2^), gender, and education were included as covariates, and students were removed.

```{r read_write_covs}
happywealth_covs <- read_rds("../results/happywealth_covs.rds")
```

@@

#### Household wealth on satisfaction (cognitive wellbeing)
```{r losat_hwnwip, fig.height=11}
library(patchwork)

p1 <- happywealth_covs %>%
  plot_linear(losat, adj_hifdip)

p2 <- happywealth_covs %>%
  plot_inflection_cubic(losat, adj_hifdip)

p3 <- happywealth_covs %>%
  plot_inflection_piecewise(losat, adj_hifdip)

(p1 + p2 + p3) +
  plot_annotation(title = "Effect of wealth on satisfaction")
```

@@

- Generally the relationship between satisfaction and household wealth appears linear
- There is no satiety point (horizontal inflection) in satisfaction before the highest decile
- Any inflection point in satisfaction is as likely to be convex as concave

#### Household wealth on happiness (affective wellbeing)
```{r gh9_hwnwip, fig.height=11}
library(patchwork)

p1 <- happywealth_covs %>%
  plot_linear(gh9, adj_hifdip)

p2 <- happywealth_covs %>%
  plot_inflection_cubic(gh9, adj_hifdip)

p3 <- happywealth_covs %>%
  plot_inflection_piecewise(gh9, adj_hifdip)

(p1 + p2 + p3) +
  plot_annotation(title = "Effect of wealth on happiness")
```

- There is no happiness satiety point (horizontal inflection) before the final decile
- A concave happiness inflection point is evident before or near the median decile

@@

as.data.frame(hifdip_gh9_fixed_fit) %>%
  select(b_alpha_Intercept, b_alpha_year2006, b_alpha_year2010, 
         b_alpha_year2014, b_alpha_year2018) %>%
  transmute(
    fy2002 = b_alpha_Intercept,
    fy2006 = mean(b_alpha_Intercept) + b_alpha_year2006,
    fy2010 = mean(b_alpha_Intercept) + b_alpha_year2010,
    fy2014 = mean(b_alpha_Intercept) + b_alpha_year2014,
    fy2018 = mean(b_alpha_Intercept) + b_alpha_year2018
  ) %>%
  mutate_all(.funs = ~brms::inv_logit_scaled(., 0, 150)*10) -> posteriors

color_scheme_set("red")
mcmc_areas(
  posteriors,
  pars = rev(colnames(posteriors)),
  prob = 0.95, # 80% intervals
  prob_outer = 1, # 99%
  point_est = "mean"
  ) +
  theme_test() +
  geom_vline(aes(xintercept = mean(posteriors$fy2002)), color = "gray", linetype = 3) +
  labs(
    subtitle = "Posterior change points (mean & 95% distribution)",
    x = "Household wealth ($000s)")
```

@@

- The change point for happiness (affective wellbeing) shifts to the right over time and occurs at greater household wealth levels.

@@

#### Change points over time

The mean change point location has moved further right (higher, faster) than the median household wealth over time:

```{r, table1, ft.align="left"}
library(knitr)

happywealth_covs %>%
  select(year, xwaveid, adj_hifdip) %>%
  mutate(
    year = paste0("fy", year),
    adj_hifdip = adj_hifdip/1000) %>%
  left_join(
    gather(
      summarise_all(posteriors, list(median)), year, change_point
    )) %>%
  group_by(year) %>%
  summarise(
    `median ($000s)` = median(adj_hifdip),
    `change point ($000s)` = mean(change_point),
    `%` = mean(adj_hifdip > mean(change_point))
  ) %>%
  kable(format = "html", caption = "Household wealth median, change point and percentile")
```

<br><br>
